Assessing \Htwo{} and \COtwo{} Network Strategies in Europe

### 2023-12-14 report/submissions/journal/paper/paper.tex
paper: rewrite section 3.3 (Hybrid model)

% alternative
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\carbongrid}{CO$_2$\=/Grid}
\newcommand{\hydrogengrid}{H$_2$\=/Grid}
\newcommand{\baelinemodel}{Baseline model}
\newcommand{\carbonmodel}{CO$_2$\=/Grid model}
\newcommand{\hydrogenmodel}{H$_2$\=/Grid model}
\newcommand{\hybridmodel}{Hybrid model}

@@

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing on the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines are already repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synfuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

@@

To our knowledge, no study has yet considered the co-optimization and comprehensive assessment of both \COtwo{} and \Htwo{} networks in a fully sector-coupled energy system. However, we argue that such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model and encompasses all relevant energy sectors. We investigate the competitive transport dynamics between \COtwo{} and \Htwo{} networks. This is done by contrasting the following four models that feature neither one of these networks, only one, or both combined:

\begin{itemize}
    \item[] \textit{Baseline:} Neither \COtwo{} nor \Htwo{} networks are constructed.
    \item[] \textit{\COgrid{}:} Only the \COtwo{} network is developed.
    \item[] \textit{\Hgrid{}:} Only the \Htwo{} network is developed.
    \item[] \textit{Hybrid:} Both \COtwo{} and \Htwo{} networks are developed.
\end{itemize}

Anticipating with tight \COtwo{} emission targets and limited sequestration assumptions, the models are exposed to the following two scenarios:
\begin{itemize}
    \item[] \textit{Net\=/Zero}: aligning with the EU's 2050 emission targets. Limiting carbon sequestration to 200~Mt/a. Also denoted as NZ.
    \item[] \textit{Net\=/Negative}: 10\% net-negative emissions relative to 1990 levels, equaling 460~Mt \COtwo{} annually. Limiting carbon sequestration to 660~Mt/a. Also denoted as NN.
\end{itemize}

We also analyze how an energy system with limited annual sequestration potential prioritizes decarbonization and fuel switching in various sectors, and how the construction of carbon and hydrogen networks varies given net-zero or net-negative emissions targets.

In the following we show how a separate investment into no or only one of the transport technologies leads to inefficient market mechanisms. We do so, by first focusing on the Net-Zero scenario and a detailed description of the different systemic implications of the model. A analysis of the .

@@

Total annual system costs in the \modBase{} are highest, totalling \label{}779 billion euros. The models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs, with \label{}3.1\% cost reduction (\label{}754~bn€) in the \modCO{}, and \label{}4.8\% (\label{}742~bn€) in the \modH{}, and \label{}5.2\% (\label{}738~bn€) in the the \modHybrid{}.

% TODO: avoid to long paranthesis
In all models, almost half of the system costs are spent on primary electricity production from wind and solar power (\label{}47\% or \label{}371~ billion euros in the \modBase{}), with another \label{}10\% (\label{}79 billion euros) spent on hydro, nuclear and biomass. Costs for energy transport networks for electricity, gas, hydrogen and carbon are below \label{}4\% (\label{}25 billion euros), with the electricity grid making up more of the system costs than the other three networks combined. This is not the case for the other models, where combined pipeline costs slightly surpass electricity transmission system costs (up to 20\% for the \hybridmodel). Capital expenditures on carbonaceous fuel production are around \label{}4\% (\label{}30 billion euros) in each model, while electrolysis makes up around \label{}4.7\% (\label{}37 billion euros). Heat pump installations account for around \label{}8\% (63 billion euros) of total system costs.

@@

By enabling the efficient transportation of carbon, hydrogen, or both, these models reduce the reliance on high-cost DAC, and source more carbon from low-cost point-source CC on biomass and gas in industrial applications (see Fig.~\ref{fig:balance_captured_carbon}). The \modH{} only captures \label{}193~Mt of \COtwo{} from DAC compared to \label{}392~Mt in the \modBase{}. Instead, the \modH{} captures more carbon from process emissions as well as gas and biomass for industry. The \modCO{} uses only \label{}154~Mt from DAC, and instead also captures more carbon than the \modH{} from gas for industry and biomass CHPs. And finally, the \modHybrid{} captures only \label{}112~Mt from DAC, but slightly more carbon from all other CC technologies.
Also, the networks reduce the reliance on methanation in the model, reducing the overall demand for CC volumes. Another knock-on effect of the reduced reliance on DAC is reduced endogenous demand for renewables to power DAC facilities.
% TODO: make sure FT and methanolisation is properly explained in methods section as almost exogenously fixed products

@@

% TODO: either explain what network models mean or use another framing
% TODO: extend first sentence
The increase in captured carbon from point sources in the network models is mostly due to CC infrastructure being deployed on industrial or power plant facilities that are present in all models, as opposed to the deployment of additional electricity generating capacity for the purpose of carbon capture. This is evident when analyzing the capture share of the different CC technologies (see Fig.~\ref{fig:captureshare_line}).

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/captureshare_line.png}
    \caption{Proportion of plants with integrated carbon capture for all models in the net zero scenario. The size of the dots corresponds to the average capacity factor of the respective technology. While peak load technologies such as gas-fired combined heat and power (CHP) plants, which are only in operation on a few days a year, are not expanded for carbon capture, baseload point sources such as biomass CHP and process emissions from industry are fully (partially) developed as soon as the transport of \COtwo{} (\Htwo{}) is permitted.}
    % A CO2 network unlock BECSS potentials, most biomass emissions are captured at point sources and transported to sequestration sites. cite~\cite{rosaAssessmentCarbonDioxide2021}
    \label{fig:captureshare_line}
\end{figure}%

In the \modBase{}, the capture share is highest for biogas-to-gas facilities, followed by process emissions, biomass for industry, biomass CHPs, gas for industry, SMR and gas CHPs. For the network models, we can observe a correlation between CC share and capacity factors of the underlying technologies: Process emissions as well as gas and biomass for industry have high capacity factors (above 80\%), and CC shares close to 100\%, the only exception being gas for industry in the \modH{}. Biomass CHPs have a capacity factor of around 50\%, and CC shares of 50\% in the \modH{} and 97\% in the \modCO{}. SMR and gas CHPs have capacity factors around 15\%, and CC shares at nearly 0\% for the \baelinemodel. The capacity factor and CC share of SMR increases to around 20\% and 50-65\%, respectively, in the other models. Gas CHPs serve as peak load electricity production technology, and with very few operating hours, thus the high investment costs into CC applications are not efficient. Similarly, SMR serves as ``peak-load'' hydrogen production technology in regions with poor renewable resources for electrolysis.

% TODO: change opener for next section: co2gird has less DAC and more CC at point sources, however h2grid is cheaper
In summary, the implementation of \COtwo{} and \Htwo{} networks is instrumental in unlocking low-cost CC potentials at geographically fixed point sources across the continent. However, despite the \modCO{} demonstrating higher absolute amounts of CC from point sources and higher CC shares across all point-source CC technologies, the \modH{} achieves lower total system costs (see Fig.~\ref{fig:cost_bar}). In the following section, we will explore the reasons for this.

@@

% TODO: smoother transition
To find the reason for why the \Htwo{} network achieves lower system costs than the \COtwo{} network, it is crucial to understand the underlying mechanisms and geographical implications of the two transportation systems. Fig.~\ref{fig:balance_map} maps the average production, consumption and flow in the carbon (top row) and hydrogen (bottom row) sectors for the \modCO{} (left) and \modH{} (right). Each map displays circles indicating the production (top half-circle) and consumption (bottom half-circle) for each region, with lines and arrows representing the average import or export volumes and flow directions between regions. The color shading of each region corresponds to the average price of the carrier.

% TODO: split the  first sentence
% TODO: add numbers: electrolyzers GW (Spain/Iberia, Germany), hydrogen prices (valleys, peaks)
As we have also discussed above, the \modCO{} captures large amounts of low-cost carbon from bioenergy and industrial activities that were untapped in the \modBase{}, especially in Central and Eastern Europe, and almost none of which is used locally (see Fig.~\ref{fig:balance_map}, top left). A large part of the captured carbon is transported from Western Europe to the Iberian Peninsula, the British Isles and Denmark, where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part of the captured carbon is transported to methanation plants in Italy and Greece, where the synthetic methane produced is fed into the well-developed gas grid, as Italy serves as a major gas import hub for Europe. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. In absence of an \Htwo{} network, the system places most electrolysis in regions with the best renewable resources to supply local CU, and smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe. The model also emphasizes gas imports into Central Europe. Hydrogen prices vary strongly across the continent, with low price ``valleys'' in regions with good renewable resources such as the Iberian Peninsula or Denmark, and high price ``peaks'' in Central Europe (see bottom left).

To understand the dynamics of the \modCO{}, we observe that the system locates CU in the regions where it can minimize costs for its material inputs, carbon and hydrogen. As the \COtwo{} network largely smoothes out carbon prices across regions, CU is located in the hydrogen price valleys on the periphery of the continent. Similarly, the model locates CS in regions where it can minimize the costs of the carbon to be sequestered. Thus, CS is located in the carbon price valleys among the subset of regions with access to offshore sequestration sites. Due to the \COtwo{} network these valleys are not strongly pronounced, but still prevalent and impact CS locations. DAC is only used for CU and thus co-located in regions with a mix of low electricity costs (for running DAC) and low hydrogen costs (where CU is located, see above).

The \Htwo{} network in the \modH{} on the other hand, transports low-cost hydrogen from regions with abundant renewable resources to regions with low-price carbon from abundant sources of fossil, biogenic and process emissions (see bottom right). The main transportation routes go from Spain and the United Kingdom to Central Europe to produce carbonaceous fuels with locally captured carbon from point sources. Also, the \Htwo{} network supplies exogenous \Htwo{} demand for industry in Central Europe at much lower costs than local electrolysis or SMR.

% TODO: implement "onshore sequestration hub" (somewhere above) for "region with access to offshore sequestration site"
To explain the dynamics of the \modH{}, we observe that the model locates CU in the carbon price valleys across Western, Central and Eastern Europe (see top right), as the \Htwo{} network smoothes out hydrogen prices across the entire continent (see bottom right). CS is again located in the carbon price valleys among those regions with access to offshore sequestration sites. Due to the lack of a \COtwo{} network the full CC potentials from point sources are exhausted in all offshore sequestration regions, and the carbon is then sequestered. The additionally needed carbon to reach 200~Mt of CS is captured via DAC in the regions with access to offshore sequestration sites with the lowest costs for DAC, with the largest share deployed in Portugal (see top right). DAC is also used for CU in this model, and for both purposes DAC is located in the regions with the lowest costs for electricity.

% TODO: write more densely (less repetitions) --> main findings: electrolysis, point-source CC even; CU different; H2 Network: co-location CU/CC; CO2 network: co-location CU/ely
In summary, both network models prefer distributed CC at point sources and electrolysis in regions with abundant renewable resources, choosing the lowest-cost production sites for each while trading off transport costs. However, differences between the two models are notable for the placement of CU, CS and DAC due to their distinct \COtwo{} and \Htwo{} price patterns. CU locations are more distributed in the \modH{}, while they cluster in a few valleys of low-cost \Htwo{} in the \modCO{}. CS co-locates in the regions with the lowest \COtwo{} costs (among those with offshore sequestration potential), but these regions are very different between the models. DAC is placed in regions with the best renewable resources in both models, but it is used for only CU in the \modCO{} and for both CU and CS in the \modH{}. Finally, the system cost advantage of the \modH{} stems from being able to supply the exogenous hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} is forced to deploy electrolysis in regions with poor renewables resources as well as additional gas infrastructure in Central Europe.

@@

% TODO: smoother transition, incl. end of 3.2
% TODO: inlcude in opener: both networks, lower system costs; motivation: what is the advantage vs both others? other advantages (more resilient)/disadvantages (not focusing on one grid)
The \modHybrid{} combines the advantages of hydrogen and carbon networks.
Fig.~\ref{fig:capacity_maps} shows the average production, consumption and transport of the two networks in the \modHybrid{}.

While the topology of the \Htwo{} network roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role.

% TODO: mention investment difference in network expansion with appendix plot (The investment in the hydrogen network is four times the investment in the carbon network.)
% TODO: rewrite: mostly short (1-2 regions) linear (no cluster/mashed network) routes towards sequestration sites (for purpose of seq.); short 1-region transports to bridge local carbon price diffs, for purpose of CU.
Similar to the \modH{}, the \modHybrid{} takes advantage of a large hydrogen network that enables hydrogen to be transported from centralized production sites in Western regions such as Spain to decentralized CU production sites close to CC point sources across the continent.
The carbon network supplements this with smaller clusters of networks. In addition to a larger \COtwo{} network cluster in Central Europe, which transports carbon to the North Sea, the system is building three large linear carbon routes. These transport carbon from Romania and Bulgaria to Greece, from northern Italy to central Italy and from Hungary, the Czech Republic and Poland to the Baltic Sea. These enable the system to transport carbon available at low cost from inland biomass sources and sequester it (see Fig.~\ref{fig:captureshare_line}). This reduces the need for DAC in the system and the sequestration sites are shifted to the end points of the carbon networks. Second-order effects can be seen in the shift in FT production compared to the \modH{}. Regions that have both good renewable resources and sequestration potential could now forego sequestration and use local process emissions for FT production, as can be observed in Ireland. A notable aspect of the hybrid model is the lack of overlap in the hydrogen and carbon network topologies. This finding is tightly linked to the fact of competing functionalities to offtake carbon emissions.
A more detailed of differences between the \modH{} and the \modHybrid{} can be found in Fig.~\ref{fig:cost_map_difference}.

% TODO: add interpretative paragraph: no CU where low-price ely, H2 is imported.
% Both networks available to smooth prices; CS draws in CC from adjacent regions; CU everywhere locally, except the CS sites; DAC centered on (very few) regions with optimal RES (and Italy/Greece for methanation)
% compare to H2 Grid model: küstennahe regionen mit longest distance to good RES(?) anzapfen, dort kein CU, CU jewtzt im Hinterland

Despite only a further reduction of \label{}0.4 percentage points (\label{}3 billion euros) in total system costs compared to the \modH{}, the \modHybrid{} brings several advantages. These include a broader range of technologies contributing to its robustness, reduced reliance on DAC, which may be costlier than anticipated in the cost projections, and less land use for wind and solar due to the decreased necessity for DAC. The increase used of biomass with carbon capture, transport and sequestration leads to slight cost advantages in comparison to the \modH{}.

@@

\ac{}In transitioning from a net zero to a net emissions reduction target, a continuation of investment trends are perceived. Fig.~\ref{fig:net-negative_cost_bar} shows the net investment changes per technology group for the \carbonmodel, the \Hgrid{} and \modHybrid{}s. The scenario imposes a net carbon removal of 460 Mt/a, with all models incurring a roughly similar net cost increase - 77 billion €/a for the \modCO{}, 81 billion €/a for the \modH{}, and 78 billion €/a for the \modHybrid{}. Both models invest relatively evenly in additional DAC, wind and heat pump systems, with DAC accounting for more than a third of the additional system costs. These remove carbon directly at sequestration sites at the coast, in particular in Portugal and UK, together with heat and power from new heat pumps at site and new solar and wind installation in the vicinity.

>>>>

% alternative
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\carbongrid}{CO$_2$\=/Grid}
\newcommand{\hydrogengrid}{H$_2$\=/Grid}
\newcommand{\baselinemodel}{Baseline model}
\newcommand{\carbonmodel}{CO$_2$\=/Grid model}
\newcommand{\hydrogenmodel}{H$_2$\=/Grid model}
\newcommand{\hybridmodel}{Hybrid model}

@@

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing on the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines are already repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

@@

To our knowledge, no study has yet considered the co-optimization and comprehensive assessment of both \COtwo{} and \Htwo{} networks in a fully sector-coupled energy system. However, we argue that such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model which optimizes operation and investments in all relevant energy sectors. To investigate the competing transport dynamics between \COtwo{} and \Htwo{} networks, we contrast the following four models that have none of these networks, only one, or both combined:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \COtwo{} nor \Htwo{} networks are constructed.
    \item[] \textit{\COgrid{}:} Only the \COtwo{} network is developed.
    \item[] \textit{\Hgrid{}:} Only the \Htwo{} network is developed.
    \item[] \textit{Hybrid:} Both \COtwo{} and \Htwo{} networks are developed.
\end{itemize}
%
In anticipation of the future \carbon{} emission targets, the models are subjected to the following two scenarios:
\begin{itemize}
    \item[] \textit{Net\=/Zero}: aligning with the EU's 2050 emission targets. Limiting carbon sequestration to 200~Mt/a. This scenario is also denoted as NZ.
    \item[] \textit{Net\=/Negative}: 10\% net-negative emissions relative to 1990 levels, equaling 460~Mt \COtwo{} annually. Limiting carbon sequestration to 660~Mt/a. Also denoted as NN.
\end{itemize}

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the models and scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. We do this by first focusing on the net-zero scenario and a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened emissions target in the Net-Negative scenario.

@@

Total annual system costs in the \modBase{} are highest, totalling \label{}779~bn€. The models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs, with \label{}3.1\% cost reduction (\label{}754~bn€) in the \modCO{}, and \label{}4.8\% (\label{}742~bn€) in the \modH{}, and \label{}5.2\% (\label{}738~bn€) in the the \modHybrid{}.

In all models, almost half of the system costs are spent on primary electricity production from wind and solar power, with around another \label{}10\% spent on hydro, nuclear and biomass. For the \baselinemodel, transmission system costs for electricity, gas, hydrogen and carbon are below \label{}4\% (\label{}25~bn€), with the electricity grid making up more of the system costs than the other three networks combined. This is not the case for the other models, where combined pipeline costs slightly surpass electricity transmission system costs (up to 20\% for the \hybridmodel). Capital expenditures on carbonaceous fuel production are around \label{}4\% (\label{}30~bn€) in each model, while electrolysis makes up around \label{}4.7\% (\label{}37~bn€). Heat pump installations account for around \label{}8\% of total system costs (64~bn€ in the \baselinemodel).

@@

By enabling the efficient transportation of carbon, hydrogen, or both, these models reduce the reliance on high-cost DAC, and source more carbon from low-cost point-source CC on biomass and gas in industrial applications (see Fig.~\ref{fig:balance_captured_carbon}). The \modH{} only captures \label{}193~Mt of \COtwo{} from DAC compared to \label{}392~Mt in the \modBase{}. Instead, the \modH{} captures more carbon from process emissions as well as gas and biomass for industry. The \modCO{} uses only \label{}154~Mt from DAC, and instead also captures more carbon than the \modH{} from gas for industry and biomass CHPs. And finally, the \modHybrid{} captures only \label{}112~Mt from DAC, but slightly more carbon from all other CC technologies.
Also, the transport of \carbon{} and \hydrogen{} reduce the reliance on methanation and therefore the overall demand for CC volumes in the model. Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

@@

The increase in captured carbon from point sources in the models with \carbon{} and/or \hydrogen{} transport is more related to an upgrading of existing industrial or power plant facilities to capture carbon rather than a shift in technologies. This is evident when analyzing the capture share of the different CC technologies (see Fig.~\ref{fig:captureshare_line}).

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/captureshare_line.png}
    \caption{Proportion of plants with integrated carbon capture for all models in the net zero scenario. The size of the dots corresponds to the average capacity factor of the respective technology. While peak load technologies such as gas-fired combined heat and power (CHP) plants, which are only in operation on a few days a year, are not expanded for carbon capture, baseload point sources such as biomass CHP and process emissions from industry are fully (partially) developed as soon as the transport of \COtwo{} (\Htwo{}) is permitted.}
    % A CO2 network unlock BECSS potentials, most biomass emissions are captured at point sources and transported to sequestration sites. cite~\cite{rosaAssessmentCarbonDioxide2021}
    \label{fig:captureshare_line}
\end{figure}%

In the \modBase{}, the capture share is highest for biogas-to-gas facilities, followed by process emissions, biomass for industry, biomass CHPs, gas for industry, SMR and gas CHPs. For models with additional transport systems, we observe a correlation between CC share and capacity factors of the underlying technologies: Process emissions as well as gas and biomass for industry have high capacity factors (above 80\%), and CC shares close to 100\%, the only exception being gas for industry in the \modH{}. Biomass CHPs have a capacity factor of around 50\%, and CC shares of 50\% in the \modH{} and 97\% in the \modCO{}. SMR and gas CHPs have capacity factors around 15\%, and CC shares at nearly 0\% for the \baselinemodel. The capacity factor and CC share of SMR increases to around 20\% and 50-65\%, respectively, in the other models. Gas CHPs serve as peak load electricity production technology, and with very few operating hours, thus the high investment costs into CC applications are not efficient. Similarly, SMR serves as ``peak-load'' hydrogen production technology in regions with poor renewable resources for electrolysis.

In summary, the implementation of \COtwo{} and \Htwo{} networks is instrumental in unlocking CC potentials and reducing system costs. However, despite the \modCO{} demonstrating higher amounts and shares of CC from point sources and therefore less DAC deployment, the \modH{} achieves even  lower total system costs (see Fig.~\ref{fig:cost_bar}).

@@

To find the reason for why the \Htwo{} network achieves lower system costs than the \COtwo{} network, it is crucial to understand the underlying mechanisms and geographical implications of the two transportation systems. Fig.~\ref{fig:balance_map} maps the average production, consumption and flow in the carbon (top row) and hydrogen (bottom row) sectors for the \modCO{} (left) and \modH{} (right). Each map displays circles indicating the production (top half-circle) and consumption (bottom half-circle) for each region, with lines and arrows representing the average import or export volumes and flow directions between regions. The color shading of each region corresponds to the average price of the carrier.

% TODO: add numbers: electrolyzers GW (Spain/Iberia, Germany), hydrogen prices (valleys, peaks)
As we have also discussed above, the \modCO{} captures large amounts of carbon from bioenergy and industrial activities that were untapped in the \modBase{}. In particular, the carbon capture takes place in Central and Eastern Europe at relatively low cost, and almost none of the carbon is used locally (see Fig.~\ref{fig:balance_map}, top left). A large part of the captured carbon is transported from Western Europe to the Iberian Peninsula (70~Mt/h), the British Isles (66~Mt/h) and Denmark (25~Mt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part of the captured carbon is transported to methanation plants in Italy and Greece (25~Mt/h), where the synthetic methane produced is fed into the well-developed gas grid, as Italy serves as a major gas import hub for Europe. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. In absence of an \Htwo{} network, the system places most electrolysis in regions with the best renewable resources to supply local CU ($\sim$1100~GW$_\text{el}$), and smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$75~GW$_\text{el}$). The model also emphasizes gas imports into Central Europe. Hydrogen prices vary strongly across the continent, with low price ``valleys'' in regions with good renewable resources such as the Iberian Peninsula or Denmark with around 60~€/MWh, and high price ``peaks'' in Central Europe with around 95~€/MWh (see bottom left).

To understand the dynamics of the \modCO{}, we observe that the system locates CU in the regions where it can minimize costs for its material inputs, carbon and hydrogen. As the \COtwo{} network largely smoothes out carbon prices across regions, CU is located in the hydrogen price valleys on the periphery of the continent. Similarly, the model locates CS in regions where it can minimize the costs of the carbon to be sequestered. Thus, CS is located in the carbon price valleys among the subset of regions with access to offshore sequestration sites. Due to the \COtwo{} network these low-price regions are not strongly pronounced, but still prevalent and impact CS locations. DAC is only used for CU and thus co-located in regions with a mix of low electricity costs (for running DAC) and low hydrogen costs.

The \Htwo{} network in the \modH{} on the other hand, transports low-cost hydrogen from regions with abundant renewable resources to regions with low-price carbon from abundant sources of fossil, biogenic and process emissions (see bottom right). The main transportation routes go from Spain (430~GW) and the United Kingdom (155~GW) to Central Europe to produce carbonaceous fuels with locally captured carbon from point sources. Also, the \Htwo{} network supplies exogenous \Htwo{} demand for industry in Central Europe (22~GW in total) at much lower costs than local electrolysis or SMR.

To explain the dynamics of the \modH{}, we observe that the model locates CU in the carbon price valleys across Western, Central and Eastern Europe (see top right), as the \Htwo{} network smoothes out hydrogen prices across the entire continent (see bottom right). CS is again located in the carbon price valleys among those regions with access to offshore sequestration sites. Due to the lack of a \COtwo{} network the full CC potentials from point sources are exhausted in all offshore sequestration regions, and the carbon is then sequestered. The additionally needed carbon to reach 200~Mt of CS is captured via DAC in the regions with access to offshore sequestration sites with the lowest costs for DAC, with the largest share deployed in Portugal (see top right). DAC is also used for CU in this model, and for both purposes DAC is located in the regions with the lowest costs for electricity.

In summary, both network models prefer CC from point sources distributed across the continent as well as electrolysis concentrated in regions with abundant renewable resources. The two network approaches reveal fundamentally different strategies to integrate these low-cost resources into the system: The \carbonmodel{} transports and sequesters low-cost carbon from distributed point sources and co-locates CU with low-cost electrolysis and DAC. The \hydrogenmodel{} distributes CU alongside with carbon point sources to which low-cost hydrogen from regions with abundant renewable resources is transported.
%
% However, differences between the two models are notable for the placement of CU, CS and DAC due to their distinct \COtwo{} and \Htwo{} price patterns. CU locations are more distributed in the \modH{}, while they cluster in a few valleys of low-cost \Htwo{} in the \modCO{}. CS co-locates in the regions with the lowest \COtwo{} costs (among those with offshore sequestration potential), but these regions are very different between the models. DAC is placed in regions with the best renewable resources in both models, but it is used for only CU in the \modCO{} and for both CU and CS in the \modH{}.
Finally, the system cost advantage of the \modH{} stems from being able to supply the exogenous hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} is forced to deploy electrolysis in regions with poor renewables resources as well as additional gas infrastructure in Central Europe.

@@

% TODO: smoother transition, incl. end of 3.2
% TODO: add network investment costs plot to Appendix. Mention investment H2 > CO2. (The investment in the hydrogen network is four times the investment in the carbon network.)
The \modHybrid{} combines the advantages of hydrogen and carbon networks. Fig.~\ref{fig:cost_bar} shows that the \modHybrid{} achieves a cost reduction of \label{}5.2\% (\label{}41 billion euros) compared to the \modBase{}, which is lower than the cost reductions of the \COgrid{} and \modH{}. Fig.~\ref{fig:capacity_maps} shows the average production, consumption and transport of carbon and hydrogen in the \modHybrid{}. While the topology of the \Htwo{} network roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role. In this section, we explore the advantages and disadvantages compared to the other models.

Similar to the \modH{}, the \modHybrid{} transports hydrogen from concentrated electrolysis sites in Western regions (Iberian Peninsula and British Isles) to distributed CU production sites across the continent. For the \COtwo{} network, this model employs short and direct pipeline routes, typically spanning one or two segments, to transport carbon from inland CC point sources to perform CS at nearby sequestration sites (e.g., Northern Germany, Italy and Greece). Additionally, the \COtwo{} network utilizes single-segment network sections to equalize local variations in carbon demand and supply to produce CU, thereby also further converging carbon prices.

To understand the dynamics of the \modHybrid{} model, we observe that the model locates CU everywere, except for a few coastal regions that perform CS, or that supply their locally captured carbon to a neigbouring sequestration region. The model uses the \COtwo{} network to transport carbon over short distances to the CS sites, and the \Htwo{} network to transport hydrogen over longer distances from the lowest-cost hydrogen regions to CU sites across the continent. DAC is deployed to supply carbon for the remaining CU needed by the model (mostly in Southern Spain).

Despite only a further reduction of \label{}0.4 percentage points (\label{}3~bn€) in total system costs compared to the \modH{}, the \modHybrid{} brings several advantages: (1) The \modHybrid{} uses both \COtwo{} and \Htwo{} networks to build a more robust energy system; (2) it relies even less on DAC than the other models, which as a technology may be costlier than anticipated in the cost projections; and (3) it requires a smaller land use footprint for wind and solar power due to lower DAC deployment. On the other hand, the \modHybrid{} needs slightly more biomass, equipped with carbon capture, to achieve its cost advantage in comparison to the \modH{}.

@@

In transitioning from a net zero to a net emissions reduction target, a continuation of investment trends are perceived. Fig.~\ref{fig:net-negative_cost_bar} shows the net investment changes per technology group for the \carbonmodel, the \Hgrid{} and \modHybrid{}s. The scenario imposes a net carbon removal of 460 Mt/a, with all models incurring a roughly similar net annual cost increase - 77~bn€ for the \modCO{}, 81~bn€ for the \modH{}, and 78~bn€ for the \modHybrid{}. Both models invest relatively evenly in additional DAC, wind and heat pump systems, with DAC accounting for more than a third of the additional system costs. These remove carbon directly at sequestration sites at the coast, in particular in Portugal and UK, together with heat and power from new heat pumps at site and new solar and wind installation in the vicinity.
